---
title: "scDHA package manual"
author: "Duc Tran"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged

editor_options:
  chunk_output_type: inline
vignette: >
  %\VignetteIndexEntry{scDHA package manual}
  %\VignetteEngine{knitr::knitr}
  %\usepackage[UTF-8]{inputenc}

---

# Introduction
Advances in microfluidics and sequencing technologies have allowed us to monitor biological systems at single-cell resolution. This comprehensive decomposition of complex tissues holds enormous potential in developmental biology and clinical research. However, the ever-increasing number of cells, technical noise, and high dropout rate pose significant computational challenges in scRNA-seq analysis.

To address this problem, we introduce single-cell Decomposition using Hierarchical Autoencoder (scDHA), which can efficiently detach noise from informative biological signals. The scDHA pipeline consists of two core modules. The first module is a non-negative kernel autoencoder that provides a non-negative, part-based representation of the data. Based on the weight distribution of the encoder, scDHA removes genes or components that have insignificant contributions to the representation. The second module is a Stacked Bayesian Self-learning Network that is built upon the Variational Autoencoder to project the data onto a low dimensional space. Using this informative and compact representation, many analyses can be performed with high accuracy and tractable time complexity (mostly linear or lower complexity). 

In one joint framework, the scDHA software package conducts cell segregation through unsupervised learning, dimension reduction and visualization, cell classification, and time-trajectory inference. 

# Installation
### Install scDHA package and other requirements
```{r, eval=FALSE}
#Install devtools: 
utils::install.packages('devtools')

#Install the package from GitHub:
devtools::install_github('duct317/scDHA')
#Or CRAN:
utils::install.packages('scDHA')

#Install tensorflow and keras in python using (install miniconda if necessary): 
if(is(try(reticulate::conda_version()), "try-error")) reticulate::install_miniconda(force = T)
keras::install_keras(method = "conda", tensorflow = "1.10.0")

#For more information about installation of keras, please visit https://keras.rstudio.com/
```

### Install other necessary packages
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("SingleCellExperiment", quietly = TRUE)) BiocManager::install("SingleCellExperiment")
if (!requireNamespace("mclust", quietly = TRUE)) install.packages("mclust")
```


# Analysis on Goolam dataset

```{r, eval=FALSE}
library(SingleCellExperiment)
library(scDHA)
#Load example data (Goolam dataset)
data("Goolam")

#Get data matrix and label
data <- t(assay(Goolam)); label <- as.character(Goolam$cell_type1)

#Log transform the data 
data <- log2(data + 1)
```

### Clustering
```{r, eval=FALSE}
#Generate clustering result
result <- scDHA(data, seed = 1)

#The clustering result can be found here 
cluster <- result$cluster

#Calculate adjusted Rand Index using mclust package
ari <- round(mclust::adjustedRandIndex(cluster,label), 2)
print(paste0("ARI = ", ari))
```

### Visualization
```{r, eval=FALSE}
#Generate 2D representation
result <- scDHA.vis(result, seed = 1)

#Plot the representation of the dataset, different colors represent different cell types
plot(result$pred, col=factor(label), xlab = "scDHA1", ylab = "scDHA2")
```

### Time trajectory inference
```{r, eval=FALSE}
#Cell stage order in Goolam dataset
cell.stages <- c("2cell", "4cell", "8cell", "16cell", "blast")

#Generate pseudo-time for each cell
result <- scDHA.pt(result, start.point = 1, seed = 1)

#Calculate R-squared value representing correlation between inferred pseudo-time and cell stage order
r2 <- round(cor(result$pt, as.numeric(factor(label, levels = cell.stages)))^2, digits = 2)

#Plot pseudo-temporal ordering of cells in Goolam dataset
plot(result$pt, factor(label, levels = cell.stages), xlab= "Pseudo Time", ylab = "Cell Stages", xaxt="n", yaxt="n")
axis(2, at=1:5,labels=cell.stages, las=2)
text(x = 1, y = 4.5, labels = paste0("R2 = ", r2))
```

### Cell classification
```{r, eval=FALSE}
#Split data into training and testing sets
set.seed(1)
idx <- sample.int(nrow(data), size = round(nrow(data)*0.75))

train.x <- data[idx, ]; train.y <- label[idx]
test.x <- data[-idx, ]; test.y <- label[-idx]

#Predict the labels of cells in testing set
prediction <- scDHA.class(train = train.x, train.label = train.y, test = test.x, seed = 1)

#Calculate accuracy of the predictions
acc <- round(sum(test.y == prediction)/length(test.y), 2)
print(paste0("Accuracy = ", acc))
```

